A Philosophy of (Inductive) Testing

### 2024-04-25 mss.tex
Final cut.

No one writes non-trivial software without testing, even if you come to believe that if it builds, it works. The first time you ran your software was a test. Chances are, you ran individual components to see if those worked. Why not take the extra few minutes to record those tests and results? With practice you will get smart about efficiently recording your throw-away tests. And after you read this I hope you will have a new appreciation of why it is worth the extra time to record your tests.

@@

Your test suite has two important non-obvious functions, as regression tests (OK, this one should be obvious) and as part of your specification. If an existing test starts failing, carefully consider Chesterton's fence \citep{WikipediaChesterton} before taking the easy road and removing the test.

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} (as 0) and \texttt{false} (as 1) \citep{Yarvin2006}, meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. In speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing, $\textrm{Succ}(0)$, gives us the first thing.  Applying $\textrm{Succ}$ to the first thing, $\textrm{Succ}(\textrm{Succ}(0))$ gives us the second thing, and so on.

@@

Through the mathematics of currying programmatic functions, we can have input arguments of multiple inductive types.\footnote{While most programming languages handle multiple function inputs via currying, Hoon gates always accept a single noun, which can be a cell.  So in this case the currying is not even theoretical.}  This results in the minimum number of tests being the number of inductive input elements squared, starting with all elements set to \lstinline[style=inlinecode]{~} and so forth.  Listing~\ref{lst:single} depicts a series of tests for the \texttt{append} gate subject to this $n^2$ explosion.

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \lstinline[style=inlinecode]{%quiz} is a well-documented Hoon implementation \citetrepo{Hjort2023}.  Once again, this solution requires some deep thinking about the code and its specification.

@@

In our experience failure modes are the most overlooked part of sofware development. It starts with passing insufficient, or no information from the programmed points of failure. So even before testing failure modes make sure you distinguish (i.e. make unique) each message from every point of failure and include any and all relevant information available for debugging (cf. Listing~\ref{lst:dup}).

@@

The standard \lstinline[style=inlinecode]{/lib/test} library, as of this writing, can test for failure but not for an expected message.  A modified testing arm (Listing~\ref{lst:error}) can be included in a testing thread to facilitate this kind of testing.  Listing~\ref{lst:urql} shows a test for an expected error message in the Obelisk database engine which uses this functionality to verify that the correct error message is raised on crash.

>>>>

No one writes non-trivial software without testing, even if you come to believe that “if it builds, it works”. The first time you ran your software was a test. Chances are, you ran individual components to see if those worked. Why not take the extra few minutes to record those tests and results? With practice you will get smart about efficiently recording your throw-away tests. And after you read this I hope you will have a new appreciation of why it is worth the extra time to record your tests.

@@

Your test suite has two important non-obvious functions, as regression tests (OK, this one should be obvious) and as part of your specification. If an existing test starts failing, carefully consider Chesterton's fence (\citeauthor{WikipediaChesterton} (2024a)) before taking the easy road and removing the test.

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} (as 0) and \texttt{false} (as 1) \citep{Yarvin2006}, meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. In speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing, $\textrm{Succ}(0)$, gives us the first thing.  Applying $\textrm{Succ}$ to the first thing, $\textrm{Succ}(\textrm{Succ}(0))$, gives us the second thing, and so on.

@@

Through the mathematics of currying programmatic functions, we can have input arguments of multiple inductive types.\footnote{While most programming languages handle multiple function inputs via currying, Hoon gates always accept a single noun, which can be a cell.  So in this case the currying is not even theoretical.}  This results in the minimum number of tests being the number of inductive input elements squared, starting with all elements set to \lstinline[style=inlinecode]{~} and so forth.  Listing~\ref{lst:single} depicts a series of tests for the \texttt{++append} gate subject to this $n^2$ explosion.

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \lstinline[style=inlinecode]{%quiz} is a well-documented Hoon implementation (\citetrepo{Hjort2023}).  Once again, this solution requires some deep thinking about the code and its specification.

@@

In our experience failure modes are the most overlooked part of sofware development. It starts with passing insufficient, or no information from the programmed points of failure. So even before testing failure modes make sure you distinguish (i.e. make unique) each message from every point of failure and include any and all relevant information available for debugging (cf.\ Listing~\ref{lst:dup}).

@@

\newpage
The standard \lstinline[style=inlinecode]{/lib/test} library, as of this writing, can test for failure but not for an expected message.  A modified testing arm (Listing~\ref{lst:error}) can be included in a testing thread to facilitate this kind of testing.  Listing~\ref{lst:urql} shows a test for an expected error message in the Obelisk database engine which uses this functionality to verify that the correct error message is raised on crash.
